Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-19 paper-new/intro.tex
add a citation in the intro
some intro work and add macros for comments

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
In modern programming language design, there is a tension between \emph{static} typing
and \emph{dynamic} typing disciplines.
With static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static 
typing, the programmer is assured that if the program passes the type-checker, their
program is free of type errors, and moreover, soundness of the equational theory implies
that program optimizations are valid.
Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their
functions. Unfortunately, most languages choose between static or dynamic typing.
As a result, programmers that initially write their code in a dynamically typed language
in order to benefit from faster prototyping and development time will need to rewrite the
entire codebase in a static language if they would like to receive the benefits of static
typing once their codebase has matured.

@@

In order for this to work as expected, gradually-typed languages should allow for
different parts of the codebase to be in different places along the spectrum from
dynamic to static, and allow for those different parts to interact with one another.
Moreover, gradually-typed languages should support the smooth migration from
dynamic typing to static typing, in that the programmer can initially leave off the
typing annotations and provide them later without altering the meaning of the program.

% Sound gradual typing
Furthermore, the parts of the program that are written in a dynamic style should soundly
interoperate with the parts that are written in a static style.
That is, the interaction between the static and dynamic components of the codebase
should preserve the guarantees made by the static types.
In particular, while statically-typed code can error at runtime in a gradually-typed language,
such an error can always be traced back to a dynamically-typed term that
violated the typing contract imposed by statically typed code.

@@

% Graduality property
Formally speaking, gradually typed languages should satisfy the 
\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
This property is also referred to as \emph{graduality} \cite{new-ahmed2018},
by analogy with parametricity.
Intuitively, graduality says that going from a dynamic to static style should not
introduce changes in the meaning of the program.
More specifically, making the types more precise by adding typing annotations
will either result in the same behavior as the original, less precise program,
or will result in a type error.

@@

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain theory}.
The techniques of synthetic guarded domain theory allow us to internalize the
step-indexed reasoning normally required in logical relations proofs of graduality,
ultimately allowing us to specify the logical relation in a manner that looks nearly
identical to a typical, non-step-indexed logical relation.

>>>>

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
In modern programming language design, there is a tension between \emph{static} typing
and \emph{dynamic} typing disciplines.
With static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static 
typing, the programmer is assured that if the program passes the type-checker, their
program is free of type errors, and moreover, soundness of the equational theory implies
that program optimizations are valid.
Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their
functions. Most languages choose between static or dynamic typing and as a result, programmers that initially write their code in a dynamically typed language
in order to benefit from faster prototyping and development time need to rewrite the
some or all of their codebase in a static language if they would like to receive the benefits of static
typing once their codebase has matured.

@@

In order for this to work as expected, gradually-typed languages should allow for
different parts of the codebase to be in different places along the spectrum from
dynamic to static, and allow for those different parts to interact with one another.
Moreover, gradually-typed languages should support the smooth migration from
dynamic typing to static typing, in that the programmer can initially leave off the
typing annotations and provide them later without altering the meaning of the program.
% Sound gradual typing
Furthermore, the parts of the program that are written in a dynamic style should soundly
interoperate with the parts that are written in a static style.
That is, the interaction between the static and dynamic components of the codebase
should preserve the guarantees made by the static types.
In particular, while statically-typed code can error at runtime in a gradually-typed language,
such an error can always be traced back to a dynamically-typed term that
violated the typing contract imposed by statically typed code.

@@

% Graduality property
One of the fundamental theorems for gradually typed languages is 
\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
and Boyland \cite{siek_et_al:LIPIcs:2015:5031}, also called \emph{graduality} \cite{new-ahmed2018},
by analogy with parametricity.
Informally, graduality says that going from a dynamic to static style should only allow for the introduction of static or dynamic type errors, and not otherwise change the meaning of the program.
%
This is a way to capture programmer intuition that increasing type
precision corresponds to a generalized form assertions at runtime
only, and so they can trust that prior observed behavior of their
dynamically typed code will remain unchanged as long as it satisfies
the new stricter typing discipline.

Additionally, gradually typed languages should offer some of the
benefits of static typing. While classical type soundness, that
well-typed programs are free from runtime errors, is not compatible
with runtime type errors, we can instead focus on \emph{type-based
reasoning}. For instance, while dynamically typed $\lambda$ calculi
only satisfy $\beta$ equality for their type formers, statically typed
$\lambda$ calculi additionally satisfy type-dependent $\eta$
properties that ensure that functions are determined by their behavior
under application and that pattern matching on data types
is safe and exhaustive.

Our goal in this work is to provide a reusable semantic framework for
gradually typed languages that can be used to prove the aforementioned
properties: graduality and type-based reasoning. \max{TODO: say more here}

@@

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
The techniques of synthetic guarded domain theory allow us to internalize the
step-indexed reasoning normally required in logical relations proofs of graduality,
ultimately allowing us to specify the logical relation in a manner that looks nearly
identical to a typical, non-step-indexed logical relation.
